---
title: "Version 3: Interactive Used Vehicle Price Estimator"
author: "Alejandro Alvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
runtime: shiny
---


<style>
body {
  background: #f5f7fb;
}

.main-container {
  max-width: 900px;
}

h1, h2 {
  color: #0B3D91;
  font-weight: 700;
}

.shiny-input-container {
  margin-bottom: 18px;
}

pre {
  background: linear-gradient(135deg, #0B3D91, #1E88E5);
  color: white;
  font-size: 18px;
  font-weight: 600;
  border-radius: 16px;
  padding: 25px;
  border: none;
  box-shadow: 0 8px 20px rgba(0,0,0,0.12);
}

.estimator-card {
  background: white;
  border-left: 6px solid #1E88E5;
  padding: 22px;
  border-radius: 16px;
  margin: 20px 0;
  box-shadow: 0 6px 18px rgba(0,0,0,0.08);
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```


```{r}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```



```{r}


library(tidyverse)
library(randomForest)
library(shiny)
library(xgboost)
library(dplyr)
```





# Interactive Used Vehicle Price Estimator


<div class="estimator-card">

Use the inputs below to estimate a used vehicle's market value using regression, Random Forest, and XGBoost models.

</div>

```{r}


cars <- read_csv("data/craigslist_cars_sample.csv")

cars_clean <- cars %>%
  mutate(
    price = as.numeric(.data$price),
    year = as.numeric(.data$year),
    mileage = str_remove_all(.data$mileage, "mi"),
    mileage = str_remove_all(mileage, ","),
    mileage = str_trim(mileage),
    mileage = case_when(
      str_detect(mileage, "k") ~ as.numeric(str_remove(mileage, "k")) * 1000,
      TRUE ~ as.numeric(mileage)
    )
  ) %>%
  filter(
    !is.na(price),
    !is.na(year),
    !is.na(mileage)
  )

valid_makes <- c(
  "Toyota", "Honda", "Ford", "BMW", "Mercedes-Benz",
  "Chevrolet", "Nissan", "Subaru", "Jeep", "Tesla",
  "Audi", "Lexus", "Dodge", "Gmc", "Hyundai", "Kia",
  "Mini", "MINI"
)

cars_v3 <- cars_clean %>%
  mutate(
    make = str_to_title(word(title, 2)),
    make = case_when(
      make == "Bmw" ~ "BMW",
      make == "Mini" ~ "MINI",
      TRUE ~ make
    )
  ) %>%
  filter(
    make %in% valid_makes,
    year >= 1990,
    year <= 2026,
    price >= 3000,
    price <= 100000,
    mileage <= 200000
  ) %>%
  dplyr::select(price, mileage, year, make)

reg_model_v3 <- lm(
  price ~ mileage + year + make,
  data = cars_v3
)

rf_model_v3 <- randomForest(
  price ~ mileage + year + make,
  data = cars_v3,
  ntree = 500
)

cars_xgb_v3 <- cars_v3 %>%
  mutate(
    make_num = as.numeric(as.factor(make))
  )

X_v3 <- as.matrix(
  cars_xgb_v3[, c("mileage", "year", "make_num")]
)

y_v3 <- cars_xgb_v3[["price"]]

xgb_model_v3 <- xgboost(
  data = X_v3,
  label = y_v3,
  nrounds = 100,
  objective = "reg:squarederror",
  verbose = 0
)

```






## Estimate a Vehicle Price


```{r}





selectInput(
  inputId = "make_input",
  label = "Vehicle Make",
  choices = sort(unique(cars_v3$make)),
  selected = sort(unique(cars_v3$make))[1]
)

numericInput(
  inputId = "year_input",
  label = "Vehicle Year",
  value = 2020,
  min = 1990,
  max = 2026
)
numericInput(
  inputId = "mileage_input",
  label = "Mileage",
  value = 75000,
  min = 0,
  max = 250000
)

verbatimTextOutput("price_output")

output$price_output <- renderText({

  new_vehicle <- data.frame(
    make = input$make_input,
    year = input$year_input,
    mileage = input$mileage_input
  )

  reg_estimate <- predict(reg_model_v3, newdata = new_vehicle)

  rf_estimate <- predict(rf_model_v3, newdata = new_vehicle)

  make_num_value <- cars_xgb_v3 %>%
    filter(make == input$make_input) %>%
    pull(make_num) %>%
    first()

  if (is.na(make_num_value)) {
    return("Please select a valid make from the dropdown menu.")
  }

  xgb_new <- as.matrix(
    data.frame(
      mileage = input$mileage_input,
      year = input$year_input,
      make_num = make_num_value
    )
  )

  xgb_estimate <- predict(xgb_model_v3, newdata = xgb_new)

  consensus_estimate <- mean(c(reg_estimate, rf_estimate, xgb_estimate))

 low_estimate <- min(c(reg_estimate, rf_estimate, xgb_estimate))
high_estimate <- max(c(reg_estimate, rf_estimate, xgb_estimate))

paste0(
  "ESTIMATED MARKET VALUE: $",
  format(round(consensus_estimate, 0), big.mark = ","),

  "\n\nPrediction Range: $",
  format(round(low_estimate, 0), big.mark = ","),
  " - $",
  format(round(high_estimate, 0), big.mark = ","),

  "\n\nModel Results",

  "\n• Regression: $",
  format(round(reg_estimate, 0), big.mark = ","),

  "\n• Random Forest: $",
  format(round(rf_estimate, 0), big.mark = ","),

  "\n• XGBoost: $",
  format(round(xgb_estimate, 0), big.mark = ",")
)
})

helpText(
  "Estimates are based on historical Craigslist listings. Accuracy may vary for makes, models, and years with limited training data."
)
```












